Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-19 figs/GSE158422_simulate.Rmd
simulation results

# parametric self-contained
simla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  dmagg1 <- agg(dm3,cores=4)
  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
  lares1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

# nonparametric self-contained
simnla <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  dmagg1 <- agg(dm3,cores=4)
  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
  lares1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
  gsig_up3 <- rownames(subset(lares1, fdr < 0.05 & t_mean > 0))
  gsig_dn3 <- rownames(subset(lares1, fdr < 0.05 & t_mean < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

# AL approach parametric self-contained
simal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # al pipeline
  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
  magg1 <- agglimma(medf1,d)
  m1 <- as.data.frame(magg1$t)
  rownames(m1) <- rownames(magg1)
  colnames(m1) <- "t"
  alres1 <- ttenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
  # summarise results
  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

# AL approach nonparametric self-contained
simnal <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # al pipeline
  medf1 <- chragg(mval=mval2,myann=myann,cores=4)
  magg1 <- agglimma(medf1,d)
  m1 <- as.data.frame(magg1$t)
  rownames(m1) <- rownames(magg1)
  colnames(m1) <- "t"
  alres1 <- wtenrich(m=m1,genesets=gsets,cores=4,testtype="selfcontained")
  # summarise results
  gsig_up3 <- rownames(subset(alres1, fdr < 0.05 & t_mean > 0))
  gsig_dn3 <- rownames(subset(alres1, fdr < 0.05 & t_mean < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(alres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

## AA Aggregate-aggregate-limma functions

Use mean value works well here.

```{r,aafunc}

gsagg <- function(x,genesets,cores=1) {
  meds <- mclapply(1:length(genesets), function(i) {
    gs = genesets[[i]]
    xx <- x[rownames(x) %in% gs,]
    med <- apply(xx,2,mean)
  },mc.cores=cores)
  mymed <- do.call(rbind,meds)
  rownames(mymed) <- names(genesets)
  as.data.frame(mymed)
}

aalimma <- function(agag,design) {
  fit.reduced <- lmFit(agag,design)
  fit.reduced <- eBayes(fit.reduced)
  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
  return(dmagg)
}

aal <- function(mval,myann,genesets,design,cores=1) {
  medf <- chragg(mval,myann,cores=cores)
  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
  aalres <- aalimma(agag=agag,design=design)
  return(aalres)
}

simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # aa pipeline
  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
  # summarise results
  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

num_dm_sets=50
sims=4

groupsizes=c(3,6,12,18)
deltas=c(0.1,0.3,0.5,0.7,1.0)
params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)
params

@@

gres

@@

gres2

@@

### parametric self-contained

```{r,runla}

lares <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

lares2 <- do.call(rbind,lapply(lares,colMeans))

lares2

@@

lacres

@@

lacres2

### nonparametric self-contained

```{r,runnla}

nlares <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simnla(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

nlares

nlares2 <- do.call(rbind,lapply(nlares,colMeans))

nlares2

@@

nlacres

@@

nlacres2

@@

### parametric self-contained

```{r,runal}

alres <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

alres

alres2 <- do.call(rbind,lapply(alres,colMeans))

alres2

@@

alcres

@@

alcres2

### nonparametric self-contained

```{r,runnal}

nalres <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simnal(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

nalres

nalres2 <- do.call(rbind,lapply(nalres,colMeans))

nalres2

@@

nalcres

@@

nalcres2

## AA sim

```{r,runaa}

aares <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

aares

aares2 <- do.call(rbind,lapply(aares,colMeans))

aares2

@@

almres

@@

almres2

>>>>

@@

@@

@@

@@

@@

num_dm_sets=50
sims=10

groupsizes=c(3,5,8,12)
deltas=c(0.1,0.2,0.3,0.4,0.5)
params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas)

params %>% kbl(caption="Parameters to test") %>% kable_paper("hover", full_width = F)

@@

@@

gres2 %>% kbl(caption="GSAmeth results") %>% kable_paper("hover", full_width = F)

gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
colnames(gres3p) <- deltas
rownames(gres3p) <- groupsizes
gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)

gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
colnames(gres3r) <- deltas
rownames(gres3r) <- groupsizes
gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)

F1(gres3p,gres3r) %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)

@@

@@

@@

lacres2 %>% kbl(caption="LA parametric results") %>% kable_paper("hover", full_width = F)

lacres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"PREC"] }))
colnames(lacres3p) <- deltas
rownames(lacres3p) <- groupsizes
lacres3p %>% kbl(caption="LA parametric precision") %>% kable_paper("hover", full_width = F)

lacres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacres2[params$groupsizes==g,"REC"] }))
colnames(lacres3r) <- deltas
rownames(lacres3r) <- groupsizes
lacres3r %>% kbl(caption="LA parametric recall") %>% kable_paper("hover", full_width = F)

F1(lacres3p,lacres3r) %>% kbl(caption="LA parametric F1") %>% kable_paper("hover", full_width = F)

@@

@@

nlacres2 %>% kbl(caption="LA nonparametric results") %>% kable_paper("hover", full_width = F)

nlacres3p <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"PREC"] }))
colnames(nlacres3p) <- deltas
rownames(nlacres3p) <- groupsizes
nlacres3p %>% kbl(caption="LA nonparametric precision") %>% kable_paper("hover", full_width = F)

nlacres3r <- do.call(rbind,lapply(groupsizes, function (g) { nlacres2[params$groupsizes==g,"REC"] }))
colnames(nlacres3r) <- deltas
rownames(nlacres3r) <- groupsizes
nlacres3r %>% kbl(caption="LA nonparametric recall") %>% kable_paper("hover", full_width = F)

F1(nlacres3p,nlacres3r) %>% kbl(caption="LA nonparametric F1") %>% kable_paper("hover", full_width = F)

@@

@@

@@

alcres2 %>% kbl(caption="AL parametric results") %>% kable_paper("hover", full_width = F)

alcres3p <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"PREC"] }))
colnames(alcres3p) <- deltas
rownames(alcres3p) <- groupsizes
alcres3p %>% kbl(caption="AL parametric precision") %>% kable_paper("hover", full_width = F)

alcres3r <- do.call(rbind,lapply(groupsizes, function (g) { alcres2[params$groupsizes==g,"REC"] }))
colnames(alcres3r) <- deltas
rownames(alcres3r) <- groupsizes
alcres3r %>% kbl(caption="AL parametric recall") %>% kable_paper("hover", full_width = F)

F1(alcres3p,alcres3r) %>% kbl(caption="AL parametric F1") %>% kable_paper("hover", full_width = F)

@@

@@

nalcres2 %>% kbl(caption="AL nonparametric results") %>% kable_paper("hover", full_width = F)

nalcres3p <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"PREC"] }))
colnames(nalcres3p) <- deltas
rownames(nalcres3p) <- groupsizes
nalcres3p %>% kbl(caption="AL nonparametric precision") %>% kable_paper("hover", full_width = F)

nalcres3r <- do.call(rbind,lapply(groupsizes, function (g) { nalcres2[params$groupsizes==g,"REC"] }))
colnames(nalcres3r) <- deltas
rownames(nalcres3r) <- groupsizes
nalcres3r %>% kbl(caption="AL nonparametric recall") %>% kable_paper("hover", full_width = F)

F1(nalcres3p,nalcres3r) %>% kbl(caption="AL nonparametric F1") %>% kable_paper("hover", full_width = F)

@@

@@

almres2 %>% kbl(caption="ALM results") %>% kable_paper("hover", full_width = F)

almres3p <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"PREC"] }))
colnames(almres3p) <- deltas
rownames(almres3p) <- groupsizes
almres3p %>% kbl(caption="ALM precision") %>% kable_paper("hover", full_width = F)

almres3r <- do.call(rbind,lapply(groupsizes, function (g) { almres2[params$groupsizes==g,"REC"] }))
colnames(almres3r) <- deltas
rownames(almres3r) <- groupsizes
almres3r %>% kbl(caption="ALM recall") %>% kable_paper("hover", full_width = F)

F1(almres3p,almres3r) %>% kbl(caption="ALM F1") %>% kable_paper("hover", full_width = F)
